\section{Related Work}
\label{sec:related}
%% The area of memory models have a long research history.
%% The starting point is the model of sequential consistency proposed by Leslie Lamport in 1979.
%% In 1979, Leslie Lamport proposed the model of sequential consistency.
%% Later
Weak memory models can be partitioned into two significantly different subclasses:
models for hardware architectures and models for programming languages.
The main difference between them is that memory models for programming languages
are expected to support compiler optimizations and may need to support compilation
to different architectures.

To this day, hardware weak memory models are relatively well-studied and understood.
%% A large amount of effort was put to formally specify models of various
%% hardware architectures. 
All major architectures have formally defined memory models:
x86~\cite{Sewell-al:CACM10},
IBM~\POWER \cite{Alglave-DAMP09,Sarkar-al:PLDI11,Alglave-al:TOPLAS14}),
\ARM~\cite{Chong-ASPLOS08, Alglave-DAMP09,Pulte-al:POPL18,Flur-al:POPL16,Alglave-al:TOPLAS14})
and \RISC~\cite{Pulte-al:POPL18}.
Among those, x86 and \POWER architectures have stable memory models which do not change in the last years,
whereas the \ARM memory model changed with transition from \ARMv{7}~\cite{Alglave-al:TOPLAS14} to \ARMv{8}~\cite{Pulte-al:POPL18}
to accommodate new instructions for shared memory access.
\RISC\footnote{https://riscv.org/} was introduced in 2010 and recently adopted a memory model~\cite{Pulte-al:POPL18}
which is almost the same as the \ARMv{8} model.

All of the models have representations in a framework of \emph{declarative} (or \emph{axiomatic})
memory models~\cite{Alglave-al:TOPLAS14}.
The framework became a standard for defining weak memory models, %(including models for programming languages),
it has tools for testing and verification~\cite{Alglave-al:TOPLAS14}.
It is also used for defining some programming language memory models: 
\CPP~\cite{Batty-al:POPL11}, \JS~\cite{Watt-al:PLDI2020}, \OCaml~\cite{Manson-al:POPL05}.
However, the framework does not allow to properly distinguish real dependencies between instructions in programs
from fake ones. For example, there is a real dependency between instructions in the snippet on the left
and a fake dependency between instructions on the right:
\[\begin{array}{l@{\qquad}|@{\qquad}l}
\readInst{}{r}{x} & \readInst{}{r}{x} \\
\writeInst{}{y}{r} & \readInst{}{y}{r * 0} \\
\end{array}\]
For hardware models, it is not a problem since they respect all \emph{syntactic} dependencies,
whereas, in the case of a programming language memory model,
if we want to support a compiler optimization which replaces $r * 0$ with $0$ in the code on the right,
we have to either distinguish them and respect only \emph{semantic} dependencies
or ignore dependencies at all (as the model of \CPP does).

\app{TODO. continue from here}

%% one might want to support  
%% For programming language models, 
%% That means 
%% has certain limitations for models of programming languages.

%% supporting compiler optimizations 
%% Alglave \etal~\cite{Alglave-al:TOPLAS14} summarized different studies in this area
%% and provided a comprehensive overview of existing models.
%% Besides that, they also proposed a general framework for specification,
%% testing and verification of hardware memory models.
%% Major architectures either have stable (x86, IBM \POWER, \ARM)
%% More recent hardware models, \eg \cite{Pulte-al:POPL18},
%% also adhere to this framework.

%% Unlike hardware memory models, programming language memory models are
%% extremely diverse, and new models appear every year.

In the context of programming language memory models the situation is more complex. 
There were various memory models proposed for different programming languages, \eg
for \Java~\cite{Manson-al:POPL05, Bender-Palsberg:OOPSLA19}, \CPP~\cite{Batty-al:POPL11}, 
\LLVM~\cite{Chakraborty-Vafeiadis:CGO17}, \JS~\cite{Watt-al:PLDI2020}, 
\OCaml~\cite{Manson-al:POPL05}, \Haskell~\cite{Vollmer-al:PPoPP17}, \etc 
However, we are unaware of any survey or detailed comparison of these models.
%The lack of such study motivated our work.

